---
title: "Exploring the health and economic imapcts of storm events in NOAA Storm Database"
author: "Eric Benzel"
date: "4/21/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis

## Loading and Processing the Raw Data
The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from the course web site: (Storm Data)[https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2] 

There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

National Weather Service Storm Data Documentation
National Climatic Data Center Storm Events FAQ
The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

```{r}

```


## Results

### Which types of events are most harmful with respect to population health in the US?

### Which types of events have the greatest economic consequences in the US?

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.
